\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-01-30 4-2-agg.Rmd
improve aggregation plot
aggregation example done

In Figure \ref{fig:basic-agg}. Aggregating them into averages or other statistics would give a better picture. This can be done with cubble through a hierarchical structure, where the cluster indicator is one level above the individual station.

The `weatherdata::climate_full` data has daily climate data of `r nrow(weatherdata::climate_full)` Australia stations from 2016 to 2020.

calcluate distance matrix
As an example to illustrate here, a kmean clustering algorithm based on the distance matrix is used and the number of centres is set to 20. More complex algorithms can also be used for more complex problem, as long as a mapping from each station id to the cluster id can be constructed.

<!-- Now we can obtain the aggregated series as described in the workflow diagram above and construct the glyph map with `GGally::glyphs()`: -->

<!-- We can also look at the precipitation of each individual station within the same cluster: -->

<!-- Lastly, there is one series on Tasmania island standing out from others, lets look at where it is: -->

```{r basic-agg, out.width="100%", out.height="90%", fig.cap="sdfasdf"}
knitr::include_graphics("figures/basic-agg.png")
```

>>>>

There are `r nrow(weatherdata::climate_full)` stations that record daily precipitation in Australia and these series will overlay on top of each other in a glyph map. Aggregation can be helpful to group series into clusters before visualising the cluster with glyph map. This example shows how to organise data at both level with the cubble function `switch_key()`.

A simple kmean algorithm based on the distance matrix is used here to create 20 clusters. This creates `station_nested` as a station level nested cubble with a cluster column indicating which group each station belongs to. More complex algorithms can be used for other problem, as long as there is a mapping from each station to a cluster.

```{r}
load(here::here("data/station_nested.rda"))
```

```{r echo = TRUE}
station_nested
```

To create a group level cubble, use `switch_key()` with the new key variable, `cluster`:

```{r echo = TRUE}
cluster_nested <- station_nested %>% switch_key(cluster) 
```

With the group level cubble, `get_centroid()` is useful to compute the centroid of each cluster, which will be used as the major axis for the glyph map later:

```{r eval = FALSE, echo = TRUE}
cluster_nested <- cluster_nested %>% get_centroid()
```

Long form cubble at both levels can be access through stretching the nested form and operations applicable to long form cubble are still available. With access to both station and cluster level cubbles, various plots can be made to understand the cluster. Figure \ref{fig:basic-agg} shows two example plots that can be made with this data: subplot A makes a glyph map with cluster level cubble and subplot B inspects the station membership of each cluster using the nested station level cubble.

```{r basic-agg, out.width="100%", fig.cap="Profile of aggregated precipitation from 639 weather stations in Australia. Subplot A shows the glyph map of the weekly averaged precipitation of each cluster. The group number of printed in the middle of y minor axis and can be used as a reference line to read the magnitude of precipitation. Subplot B shows the station membership of each cluster"}
knitr::include_graphics("figures/basic-agg.png")
```

### 2022-01-30 main.tex
aggregation example done

With recent climate, there are 639 stations in Australia that records
daily maximum and minimum temperature and precipitation. While it would
be too much series to show on a glyph map, stations close to each other
should have similar precipitation pattern. Aggregation can be helpful
here to cluster series into groups and visualise them with glyph map.

A simple kmean algorithm is used based on the distance matrix to cluster
the stations into 20 groups. This creates \texttt{station\_nested} as a
station level nested cubble with a cluster column indicating which group
each station belongs to. More complex algorithms can be used for more
complex problem, as long as a mapping from each station id to the
cluster id can be constructed.

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [639]: nested form
# bbox:     [113.53, -43.66, 153.64, -10.05]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name             wmo_id ts               cluster
  <chr>       <dbl> <dbl> <dbl> <chr>             <dbl> <list>             <int>
1 ASN00001006 -15.5  128.   3.8 wyndham aero      95214 <tibble [1,827 ~       3
2 ASN00001007 -13.8  126.   6   troughton island  94102 <tibble [1,827 ~       3
3 ASN00001018 -16.4  126. 546   mount elizabeth   94211 <tibble [1,460 ~      16
4 ASN00001019 -14.3  127.  23   kalumburu         94100 <tibble [1,827 ~       3
5 ASN00001020 -14.1  126.  51   truscott          95101 <tibble [1,827 ~       3
# ... with 634 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

With the group level cubble, it is useful to compute the centroid of
each cluster as the location of group level glyph map.
\texttt{get\_centroid()} is a shortcut for doing so or you can manually
find the convex hull, its centroid, and extract the longitude and
latitude:

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_nested <- cluster_nested %>% 
+   get_centroid()
\end{CodeInput}
\end{CodeChunk}

Long form cubble at both levels can be access through stretching the
nested form. Here weekly averages of the precipitation is calculated:

\begin{CodeChunk}
\begin{CodeInput}
R> station_long <- station_nested %>% 
+   stretch(ts) %>% 
+   mutate(wk = lubridate::week(date)) %>% 
+   group_by(wk) %>% 
+   summarise(prcp = sum(prcp, na.rm = TRUE))
R> 
R> cluster_long <- cluster_nested %>% 
+   stretch(ts) %>% 
+   mutate(wk = lubridate::week(date)) %>% 
+   group_by(wk) %>% 
+   summarise(prcp = mean(prcp, na.rm = TRUE)) %>% 
+   migrate(cent_long, cent_lat)
\end{CodeInput}
\end{CodeChunk}

With the four cubbles: \texttt{station\_nested}, \texttt{station\_long},
\texttt{cluster\_nested}, and \texttt{cluster\_long}, all the
information at both levels are accessible. This allows us to make glyph
map on the cluster level, looking at station composition of each
cluster, and plot the individual series to check for consistency within
groups. Figure \ref{fig:basic-agg} composes of the three plots together
to show the precipitation in Australia.

@@

\caption[Aggregated precipitation in Australia]{Aggregated precipitation in Australia.}\label{fig:basic-agg}
\end{figure}
\end{CodeChunk}

>>>>

There are 639 stations that record daily precipitation in Australia and
these series will overlay on top of each other in a glyph map.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map. This example shows how to
organise data at both level with the cubble function
\texttt{switch\_key()}.

A simple kmean algorithm based on the distance matrix is used here to
create 20 clusters. This creates \texttt{station\_nested} as a station
level nested cubble with a cluster column indicating which group each
station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

\begin{CodeChunk}
\begin{CodeInput}
R> station_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [639]: nested form
# bbox:     [113.53, -43.66, 153.64, -10.05]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name             wmo_id ts               cluster
  <chr>       <dbl> <dbl> <dbl> <chr>             <dbl> <list>             <int>
1 ASN00001006 -15.5  128.   3.8 wyndham aero      95214 <tibble [1,827 ~       3
2 ASN00001007 -13.8  126.   6   troughton island  94102 <tibble [1,827 ~       3
3 ASN00001018 -16.4  126. 546   mount elizabeth   94211 <tibble [1,460 ~      16
4 ASN00001019 -14.3  127.  23   kalumburu         94100 <tibble [1,827 ~       3
5 ASN00001020 -14.1  126.  51   truscott          95101 <tibble [1,827 ~       3
# ... with 634 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

With the group level cubble, \texttt{get\_centroid()} is useful to
compute the centroid of each cluster, which will be used as the major
axis for the glyph map later:

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_nested <- cluster_nested %>% get_centroid()
\end{CodeInput}
\end{CodeChunk}

Long form cubble at both levels can be access through stretching the
nested form and operations applicable to long form cubble are still
available. With access to both station and cluster level cubbles,
various plots can be made to understand the cluster. Figure
\ref{fig:basic-agg} shows two example plots that can be made with this
data: subplot A makes a glyph map with cluster level cubble and subplot
B inspects the station membership of each cluster using the nested
station level cubble.

@@

\caption[Profile of aggregated precipitation from 639 weather stations in Australia]{Profile of aggregated precipitation from 639 weather stations in Australia. Subplot A shows the glyph map of the weekly averaged precipitation of each cluster. The group number of printed in the middle of y minor axis and can be used as a reference line to read the magnitude of precipitation. Subplot B shows the station membership of each cluster}\label{fig:basic-agg}
\end{figure}
\end{CodeChunk}
